### 2016-01-13 paper/pldi16-submission/response.txt

We thank the reviewers for their enthusiastic comments, and will briefly
respond to the main critiques.

Formalism
---------

While we have implemented (and evaluated) our technique on the pure
subset of OCaml -- including algebraic datatypes, polymorphism, and type
annotations -- we restricted the formalism to a simpler language because
the additional constructs do not any major complexity to the technique.
We simply treat constructor application, case analysis, type annotation,
etc. as primitive operations like `+` and `if`; they narrow the partial
types of values and force instantiation. However, in light of the
reviews it is clear that this is not obvious to the reader, so we will
include an extended formal semantics in the next iteration of the paper.

We also thank Reviewer B in particular for pointing out the two bugs in
our proof. We have refactored the substitutions in the formalism to be
properly threaded through `narrow` and `E-Plus-Good`, and have changed
our definition of type compatibility to the usual one. While partial
types do grow monotonically, in Lemma 4 specifically we are concerned
with *any* value (and type) so the example with incompatible pair types
applies.

Representivity of the datasets
------------------------------

Between our dataset and the UW dataset, we have numerous instances of
writing higher-order functions and instantiating their arguments, as
well as two different expression evaluators. So we believe the
benchmarks are representative of the range of programs an undergraduate
student may have to write.

Pedagogic motivation for witnesses and traces
---------------------------------------------

We are also convinced of the importance of witnesses and traces for
debugging and did not intend this paper to specifically present evidence
in their favor, rather to show that generating witnesses is feasible and
*practical* for type errors. (This is also why we did not discuss
producing witnesses for other runtime errors like divide-by-zero.) We
will update the motivation to be clearer about the scope of the paper
and results.

### 2016-01-13 paper/pldi16-submission/response.txt

Representivity of the datasets
------------------------------

Between our dataset and the UW dataset, we have numerous instances of
writing higher-order functions and instantiating their arguments, as
well as two different expression evaluators. So we believe the
benchmarks are representative of the range of programs an undergraduate
student may have to write.

Pedagogic motivation for witnesses and traces
---------------------------------------------

We are also convinced of the importance of witnesses and traces for
debugging and did not intend this paper to specifically present evidence
in their favor, rather to show that generating witnesses is feasible and
practical for type errors. (This is also why we did not discuss
producing witnesses for other runtime errors like divide-by-zero.) We
will update the motivation to be clearer about the scope of the paper
and results.

### 2016-01-20 paper/icfp16-submission/response.txt

We thank the reviewers for their enthusiastic comments, and will briefly
respond to the main critiques.

Formalism
---------

While we have implemented (and evaluated) our technique on a pure subset
of OCaml -- including algebraic datatypes, polymorphism, and type
annotations -- we restricted the formalism to a simpler language because
the additional constructs do not add significant complexity to the
technique. We simply treat constructor application, case analysis, type
annotation, etc. as primitive operations like `+` and `if`; they narrow
the partial types of values and force instantiation. However, in light
of the reviews it is clear that this is not obvious to the reader, so we
will add text to Section 3 to list additional language features that can
be handled as primitive operations from the perspective of our
semantics.

We also thank Reviewer B in particular for pointing out two bugs in our
proof. We have refactored the substitutions in the formalism to be
properly threaded through `narrow` and `E-Plus-Good`. We have also
changed our definition of type compatibility to the usual one, fixing the

  (([]_1, []_2), []_3) ~ ([]_4, ([]_5, []_6))

example (while partial types do grow monotonically, in Lemma 4
specifically we are concerned with *any* value, so we must account for
this case).

Representivity of the datasets
------------------------------

Between our dataset and the UW dataset, we have numerous instances of
writing higher-order functions and instantiating their arguments, as
well as two different expression evaluators. We will expand the text in
Section 5 to describe the generality of the datasets, including a list
of language features and idioms that are present.

Pedagogic motivation for witnesses and traces
---------------------------------------------

We are also convinced of the importance of witnesses and traces for
debugging and did not intend this paper to specifically present evidence
in their favor. We will update the motivation and conclusion to clarify
that the scope of the paper is to present algorithmic and empirical
evidence that generating witnesses is practical for type errors.

### 2016-01-20 paper/icfp16-submission/abstract.tex
copy to icfp16-submission, switch to icfp class

\begin{abstract}
  %
  \emph{Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a \emph{dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate \emph{how} an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we first symbolically
  execute the body to dynamically synthesize witness values
  that can make the program go wrong.
  We prove that our procedure synthesizes
  \emph{general witnesses} in that if a witness is
  found, then \emph{for all} input types, there exist
  inhabitants that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a \emph{reduction graph} that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate our approach on two data sets
  comprising over 4,500 ill-typed student programs.
  Our technique is able to generate witnesses for
  88\% of the programs, and our reduction graph
  is yields small counterexamples for 90\% of the witnesses.
\end{abstract}

>>>>

\begin{abstract}
  %
  \emph{Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a \emph{dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate \emph{how} an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we symbolically
  execute the body to dynamically synthesize witness values
  that can make the program go wrong.
  We prove that our procedure synthesizes
  \emph{general witnesses} in that if a witness is
  found, then \emph{for all} input types, there exist
  inhabitants that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a \emph{reduction graph} that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate our approach on two data sets
  comprising over 4,500 ill-typed student programs.
  Our technique is able to generate witnesses for
  88\% of the programs, and our reduction graph
  yields small counterexamples for 90\% of the witnesses.
\end{abstract}
